---
subtitle: "Pooled Kaplan-Meier estimates after multiple imputation and propensity score analysis"
author: Janick Weberpals, RPh, PhD
date: last-modified
format: html
toc: true
toc-depth: 3
keep-md: true
embed-resources: true
editor: visual
bibliography: ../references.bib
csl: ../pharmacoepidemiology-and-drug-safety.csl
---

# Pooled Kaplan-Meier Estimates {#sec-pooled-km-estimates}

In this vignette, we demonstrate how to use the `km_pooling()` function from the `encore.analytics` package to compute and visualize pooled Kaplan-Meier estimates across multiply imputed and propensity score matched/weighted datasets.

## Overview

The `km_pooling()` function addresses a common challenge in survival analysis with missing data: how to properly combine Kaplan-Meier estimates across multiple imputed datasets after propensity score matching or weighting. The function implements the methodology recommended by @marshall2009 for pooling survival probabilities using Rubin's rules.

## Key Features

The function provides three main outputs:

1.  Pooled median survival estimates with confidence intervals
2.  A detailed survival probability table with pooled estimates
3.  A visualization of the pooled Kaplan-Meier curve

## Methodology

The function follows these key steps:

1.  Fits Kaplan-Meier survival functions to each imputed and matched/weighted dataset
2.  Transforms survival probabilities using complementary log-log transformation
3.  Pools transformed survival probabilities and computes total variance using Rubin's rule
4.  Back-transforms pooled survival probabilities and computes 95% confidence intervals
5.  Extracts median survival time and corresponding 95% confidence intervals
6.  Plots the Kaplan-Meier curve with pooled survival probabilities and confidence intervals

## Mathematical Details

The pooling process involves several mathematical transformations to ensure proper combination of survival estimates across imputed datasets:

### 1. Complementary Log-Log Transformation

For each time point $t$ and imputed dataset $m$, the survival probability $S_m(t)$ is transformed using the complementary log-log transformation:

$$g_m(t) = \log(-\log(1-S_m(t)))$$

This transformation helps approximate normality, which is required for proper application of Rubin's rules.

### 2. Pooling Transformed Estimates

The pooled estimate $\bar{Q}$ (following Rubin's notation) at each time point is computed as:

$$\bar{Q} = \frac{1}{M}\sum_{m=1}^M g_m(t)$$

where $M$ is the number of imputed datasets.

### 3. Variance Estimation

The total variance $T$ of the pooled estimate combines within-imputation variance $\bar{U}$ and between-imputation variance $B$:

$$\bar{U} = \frac{1}{M}\sum_{m=1}^M U_m$$

$$B = \frac{1}{M-1}\sum_{m=1}^M (g_m(t) - \bar{Q})^2$$

$$T = \bar{U} + (1 + \frac{1}{M})B$$

where $U_m$ is the variance of $g_m(t)$ in the $m$th imputed dataset.

### 4. Back-Transformation

The pooled survival probability and its confidence intervals are obtained by back-transforming:

$$S(t) = 1 - \exp(-\exp(\bar{Q}))$$

The 95% confidence intervals are computed as:

$$S_{lower}(t) = 1 - \exp(-\exp(\bar{Q} - 1.96\sqrt{T}))$$ $$S_{upper}(t) = 1 - \exp(-\exp(\bar{Q} + 1.96\sqrt{T}))$$

## Example Application

Let's walk through a complete example using simulated data:

```{r}
#| label: setup
#| message: false

library(here)
library(dplyr)
library(survival)
library(mice)
library(MatchThem)
library(encore.analytics)

source(here("functions", "covariate_vectors.R"))
```

### Data Generation

First, we'll simulate a dataset with some missing values:

{{< include _snippets/_data_generation.qmd >}}

### Multiple Imputation

We'll create 10 imputed datasets:

{{< include _snippets/_imputation.qmd >}}

### Propensity Score Weighting

We'll apply propensity score matching to each imputed dataset:

{{< include _snippets/_ps_fit.qmd >}}

{{< include _snippets/_matching.qmd >}}

### Computing Pooled Kaplan-Meier Estimates

Now we can use `km_pooling()` to compute and visualize the pooled survival curves:

```{r}
#| label: km_pooling

# specify the survival model
km_fit <- as.formula(Surv(fu_itt_months, death_itt) ~ treat)

# estimate and pool median survival times and Kaplan-Meier curve
km_out <- km_pooling(
  x = mimids_data,
  surv_formula = km_fit
  )

# View median survival time estimates
km_out$km_median_survival

# Plot the Kaplan-Meier curve
km_out$km_plot
```

## Interpretation

The output provides:

1.  **Median Survival Times (`km_median_survival`):** Pooled median survival estimates and 95% confidence intervals for each treatment group.

-   strata = stratum
-   t_median = median survival time
-   t_lower = lower 95% CI of median survival time
-   t_upper = upper 95% CI of median survival time

2.  **Kaplan-Meier survival table (`km_survival_table`):**

-   strata = stratum

-   time = observed time point

-   m = number of imputed datasets

-   qbar = pooled univariate estimate of the complementary log-log transformed survival probabilities, see formula (3.1.2) Rubin (1987)

-   t = total variance of the pooled univariate estimate of the complementary log-log transformed survival probabilities, formula (3.1.5) Rubin (1987)

-   se = total standard error of the pooled estimate (derived as sqrt(t))

-   surv = back-transformed pooled survival probability

-   lower = Wald-type lower 95% confidence interval of back-transformed pooled survival probability

-   upper = Wald-type upper 95% confidence interval of back-transformed pooled survival probability

3.  **Kaplan-Meier Curve (**`km_plot`**)**: A visualization showing the pooled survival probabilities over time with confidence bands (ggplot2 object)

## Technical Details

The function handles several nuances in survival analysis:

-   Transforms survival probabilities to approximate normality using complementary log-log transformation
-   Applies Rubin's rules for pooling estimates and computing variance
-   Handles edge cases in median survival time estimation
-   Incorporates weights and cluster membership for matched/weighted datasets

## References
